Statistically-estimated tree composition for the northeastern United States \\ at the time of Euro-American settlement

### 2016-02-03 paperPlos.tex
revised and added for plos revision january 2016

@@

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Statistically-estimated Tree Composition for the\\
Northeastern United States at Euro-American Settlement}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Christopher J. Paciorek\textsuperscript{1,*},
Simon J. Goring\textsuperscript{2,\Yinyang},
Andrew L. Thurman\textsuperscript{3,\Yinyang},
Charles V. Cogbill\textsuperscript{4},
John W. Williams\textsuperscript{2,5},
David J. Mladenoff\textsuperscript{6},
Jody A. Peters\textsuperscript{7},
Jun Zhu\textsuperscript{8},
Jason S. McLachlan\textsuperscript{7,\ddag}
\\
\bigskip
\bf{1} Department of Statistics, University of California, Berkeley, California,
USA
\\
\bf{2} Department of Geography, University of Wisconsin, Madison, Wisconsin,
USA
\\
\bf{3} Department of Statistics, University of Iowa, Iowa City, Iowa, USA
\\
\bf{4} Harvard Forest, Harvard University, Petersham, Massachusetts, USA
\\
\bf{5} Center for Climatic Research, University of Wisconsin, Madison,
Wisconsin, USA
\\
\bf{6} Department of Forestry and Wildlife Ecology, University of Wisconsin,
Madison, Wisconsin, USA
\\
\bf{7} Department of Biological Sciences, University of Notre Dame, Notre
Dame, Indiana, USA
\\
\bf{8} Department of Statistics, University of Wisconsin, Madison, Wisconsin,
USA
\\
\bigskip

@@

We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on the
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-sized
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4.

@@

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \cite{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \cite{friedman2005regional,rhemtulla2009historical}
and with various spatial resolutions, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \cite{goring2015composition}
provide a new dataset of forest composition, biomass, and stem density
based on PLS data for the upper Midwest that is resolved to an 8 km
by 8 km grid cell scale, providing broad spatial coverage at a spatial
scale that can be compared to modern forests using Forest Inventory
and Analysis products \cite{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
data from southern Michigan, and with the TPS data, this gives us
raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data only provide estimates of
within-cell variance that do not account for information from nearby
locations. Second, there are data gaps: the available digitized data
from Illinois and Indiana represent a small fraction of those states
and missing townships are common in the TPS data. Third, the TPS and
PLS data have fundamentally different sampling design and spatial
resolution. Our statistical model allows us to provide a spatially-complete
data product of settlement-era tree composition for a common 8 km
grid with uncertainty across the northeastern U.S.

In \emph{Data} we describe the data sources, while in \emph{Statistical model}
we describe the statistical model used to create the data product.
In \emph{Model comparison} we quantitatively compare competing
statistical specifications, and \emph{Data Product}
we describe the final data product. In \emph{Discussion}
we discuss limitations of the statistical model, and we list related
data products under development.

@@

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. There are also data from a single township
in Quebec and a single township in northern Delaware. Digitization
of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
with PLS data for nearly all 8 km grid cells, but data in Illinois
and Indiana represent a sample of the full set of grid cells, with
survey record transcription ongoing. Data for the eastern states are
available for a subset of the full set of townships covering the domain;
the TPS data for some townships were lost, incomplete, or have not
been located \cite{Cogb:etal:2002}.

@@

\caption{{\bf Spatial domain of the northeastern United States, with locations with
data shown in gray.} \\ Locations are grid cells in midwestern portion
and townships in eastern portion. In addition to locations without
data being indicated in white, grid cells completely covered in water
are white (e.g., a few locations in the northwestern portion of the
domain in the states of Minnesota and Wisconsin).}}
\label{fig:Spatial-domain}
\end{figure}

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the upper Midwest
are available through the NIS Data portal (accession number and version
to be inserted when available). The TPS data were compiled by C.V.
Cogbill from a myriad of archival sources representing land division
surveys conducted in connection with local settlement.

The aggregation into taxonomic groups is primarily at the genus level
but is at the species level in some cases of monospecific genera.
We model the following 22 taxa plus an ``other hardwood'' category:
Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash (\emph{Fraxinus
spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus grandifolia)},
Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa sylvatica}
and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
Elm (\emph{Ulmus spp.}), Fir (\emph{Abies balsamea}), Hemlock (\emph{Tsuga
canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
caroliniana} and \emph{Ostrya virginiana}), Maple (\emph{Acer spp.}),
Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}), Poplar/tulip
poplar (\emph{Populus spp.}~and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
Walnut (\emph{Juglans spp.}). Note that in several cases (black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different genera or even families. For the midwestern subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these species have 0 and 7 trees present, respectively. The taxa
grouped into the other hardwood category are those for which fewer
than roughly 2000 trees were present in the dataset; however we include
Atlantic white cedar explicitly despite it only having 336 trees in
the dataset because of specific ecological interest in Atlantic white
cedar wetlands.

@@

Our domain is a rectangle covering all of the states using a metric
Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
with the rectangle split into 8 km cells, arranged in a 296 by 180
grid of cells, with the centroid of the cell in the southwest corner
located at (-71000 m, 58000 m). For the midwestern subdomain we use
the western-most 146 by 180 grid of cells. For the eastern subdomain
we use the eastern-most 180 by 180 grid of cells and then omit 23
rows of cells in the north and 17 rows of cells in the south, as these
grid cells are outside of the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in grid cell or township.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section) and can be used in subsequent analyses. The
mean and standard deviation of the samples for each pair of cell and
taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of $P$ Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the $m$ grid cells. In \emph{Latent variable model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. This process is a prior
structure that serves to smooth across noise in the observations and
allows for interpolation to locations with no data. Apart from the
sum-to-one constraint, the taxa are considered to be independent in
the prior. We did not want to impose any structure that ties the different
taxa together, as any correlation will likely vary across space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \cite{Lind:etal:2011}.

@@

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}$
is computationally simple \cite{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation helps to alleviate this problem.
Next we describe two alternative spatial models that we considered;
in \emph{Model comparison}, we use cross-validation to
choose between these two representations.

@@

Our first model is a standard conditional autoregressive (CAR) model
\cite{Bane:etal:2004}. We use a standard form of this model, which
treats the four cardinal neighbors of each grid cell as the neighbors
of the grid cell. The corresponding precision matrix has diagonal
elements, $Q_{ii}$, equal to the number of neighbors for the \emph{i}th
area (i.e., four except for cells on the boundary of the domain),
while $Q_{ik}=-1$ (the negative of a weight of one) when areas \emph{i}
and \emph{k} are neighbors and $Q_{ik}=0$ when they are not. This gives
the following model for the values of $\alpha_{p}(s_{i})$ collected
as a vector across all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\N(0,\sigma_{p}^{2}Q^{-}).
\]
The use of the generalized inverse notation indicates that \emph{Q} is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where $C$ is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$C{}_{ik}=1$ if locations \emph{i} and \emph{k} are neighbors and zero otherwise.
The matrix \emph{D} is an $m\times m$ diagonal matrix containing the
row sums of matrix \emph{C} as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$

@@

Gaussian processes (GP) are also standard models for spatial processes \cite{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of manipulations involving large covariance matrices. Given this,
\cite{Lind:etal:2011} proposed a new framework for using Gaussian
MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
partial differential equations (SPDEs).

We consider Gaussian processes in the \matern class, using the following
parameterization of the \matern correlation function, 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),
\end{equation}
where \emph{d} is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
the distance between the locations. Considering all pairs of locations,
this defines a correlation matrix for all locations of interest.

The  approach of \cite{Lind:etal:2011} allows us to consider MRF approximations
to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the \emph{Q} matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
neighbors are -2. Those for diagonal neighbors are 2, and those
for 2nd-order cardinal neighbors are 1. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

The primary difference between the CAR and Lindgren models is that
the Lindgren model provides an additional degree of freedom by estimating
$\rho$. In particular $\rho$ allows us to estimate the locality
of the smoothing. As $\rho$ decreases, the model uses increasingly
localized data to estimate the compositional proportions at a given
location, effectively averaging the empirical proportions over smaller
neighborhoods. In general, the model of \cite{Lind:etal:2011} will generally
provide for a smoother estimate than the CAR model \cite{Paci:2013}.

To ensure that the $\sigma^{2}$ parameter is mathematically equivalent
between the two models, we reparameterize, producing our second model:
\[
\alpha_{p}\sim\N\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
\]

@@

Here we summarize the results of our cross-validation analyses that
inform the choice between the CAR and SPDE models.

>>>>

\usepackage{color}

@@

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Statistically-estimated Tree Composition for the\\
Northeastern United States at Euro-American Settlement}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Christopher J. Paciorek\textsuperscript{1,*},
Simon J. Goring\textsuperscript{2,\Yinyang},
Andrew L. Thurman\textsuperscript{3,\Yinyang},
Charles V. Cogbill\textsuperscript{4},
John W. Williams\textsuperscript{2,5},
David J. Mladenoff\textsuperscript{6},
Jody A. Peters\textsuperscript{7},
Jun Zhu\textsuperscript{8},
Jason S. McLachlan\textsuperscript{7,\ddag}
\\
\bigskip
\bf{1} Department of Statistics, University of California, Berkeley, California,
USA
\\
\bf{2} Department of Geography, University of Wisconsin, Madison, Wisconsin,
USA
\\
\bf{3} Department of Statistics, University of Iowa, Iowa City, Iowa, USA
\\
\bf{4} Harvard Forest, Harvard University, Petersham, Massachusetts, USA
\\
\bf{5} Center for Climatic Research, University of Wisconsin, Madison,
Wisconsin, USA
\\
\bf{6} Department of Forest and Wildlife Ecology, University of Wisconsin,
Madison, Wisconsin, USA
\\
\bf{7} Department of Biological Sciences, University of Notre Dame, Notre
Dame, Indiana, USA
\\
\bf{8} Department of Statistics, University of Wisconsin, Madison, Wisconsin,
USA
\\
\bigskip

@@

We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on the
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-shaped
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4.

@@

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \cite{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \cite{friedman2005regional,rhemtulla2009historical}
and with various spatial resolutions, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \cite{goring2015composition}
provide a new dataset of forest composition, biomass, and stem density
based on PLS data for the upper Midwest that is resolved to an 8 km
by 8 km grid cell scale, providing broad spatial coverage at a spatial
scale that can be compared to modern forests using Forest Inventory
and Analysis products \cite{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
data from southern Michigan, and with the TPS data, this gives us
raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data only provide estimates of
within-cell variance that do not account for information from nearby
locations. Second, there are data gaps: the available digitized data
from Illinois and Indiana represent a small fraction of those states,
and missing townships are common in the TPS data. Third, the TPS and
PLS data have fundamentally different sampling design and spatial
resolution. Our statistical model allows us to provide a spatially-complete
data product of settlement-era tree composition for a common 8 km
grid with uncertainty across the northeastern U.S.

In \emph{Data} we describe the data sources, while in \emph{Statistical model} we describe 
the statistical model used to create the data product. In In \emph{Model comparison}
 we quantitatively compare competing
statistical specifications, and in \emph{Data Product}
we describe the final data product. In \emph{Discussion}
we discuss the uncertainties estimated by and the
limitations of the statistical model, and we list related data products
under development.

@@

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid, but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. We make predictions for all
of the states listed above; these constitute our core domain. There
are also data from a single township in Quebec and a single township
in northern Delaware; these data help inform predictions
in nearby locations within our core domain, but predictions are not
made for Quebec or Delaware. Digitization of PLS data in Minnesota,
Wisconsin and Michigan is essentially complete, with PLS data for
nearly all 8 km grid cells, but data in Illinois and Indiana represent
a sample of the full set of grid cells, with survey record transcription
ongoing. Data for the eastern states are available for a subset of
the full set of townships covering the domain; the TPS data for some
townships were lost, incomplete, or have not been located \cite{Cogb:etal:2002}.

@@

\caption{{\bf Spatial domain of the northeastern United States, with locations with
data shown in gray.} \\ Locations are grid cells in midwestern portion
and townships in eastern portion. In addition to locations without
data being indicated in white, grid cells completely covered in water
are white (e.g., a few locations in the northwestern portion of the
domain in the states of Minnesota and Wisconsin).}
\label{fig:Spatial-domain}
\end{figure}

@@

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the midwest
subdomain are available through the NIS data portal as
DOI:10.5072/FK2/e9c144ab83af70e29b8d6647aaae58f5 \cite{Gori:etal:data:2016}.
The TPS data were compiled by C.V. Cogbill from a myriad of archival
sources representing land division surveys conducted in connection
with local settlement and are available through the
NIS data portal as DOI:10.5072/FK2/4d44aebe71dbf65290d8ce1bde03598b
and DOI:10.5072/FK2/88e167487b22d5f657e7ce7e033772e1 \cite{Cogb:dataNE:2016,Cogb:dataOH:2016}.

The aggregation into taxonomic groups is primarily at the genus level
but is at the species level in some cases of monospecific genera.
We model the following 22 taxa plus an ``other hardwood'' category:
Atlantic white cedar (\emph{Chamaecyparis thyoides}), Ash (\emph{Fraxinus
spp.}), Basswood (\emph{Tilia americana}), Beech (\emph{Fagus grandifolia)},
Birch (\emph{Betula spp.}), Black gum/sweet gum (\emph{Nyssa sylvatica}
and \emph{Liquidambar styraciflua}), Cedar/juniper (\emph{Juniperus
virginiana} and \emph{Thuja occidentalis}), Cherry (\emph{Prunus spp.}),
Chestnut (\emph{Castanea dentata}), Dogwood (\emph{Cornus spp.}),
Elm (\emph{Ulmus spp.}), Fir (\emph{Abies balsamea}), Hemlock (\emph{Tsuga
canadensis}), Hickory (\emph{Carya spp.}), Ironwood (\emph{Carpinus
caroliniana} and \emph{Ostrya virginiana}), Maple (\emph{Acer spp.}),
Oak (\emph{Quercus spp.}), Pine (\emph{Pinus spp.}), Poplar/tulip
poplar (\emph{Populus spp.}~and \emph{Liriodendron tulipifera}),
Spruce (\emph{Picea spp.}), Tamarack (\emph{Larix laricina}), and
Walnut (\emph{Juglans spp.}). Note that in several cases (black gum/sweet
gum, ironwood, poplar/tulip poplar, cedar/juniper), because of ambiguity
in the common tree names used by surveyors, a group represents trees
from different genera or even families. For the midwestern subdomain
we do not fit statistical models for Atlantic white cedar and chestnut
as these species have 0 and 7 trees present, respectively. The taxa
grouped into the other hardwood category are those for which fewer
than roughly 2000 trees were present in the dataset; however, we include
Atlantic white cedar explicitly despite it only having 336 trees in
the dataset because of specific ecological interest in Atlantic white
cedar wetlands.

@@

Our domain is a rectangle covering all of the states using a metric
Albers (Great Lakes and St. Lawrence) projection (PROJ4: EPSG:3175),
with the rectangle split into 8 km cells, arranged in a 296 by 180
grid of cells, with the centroid of the cell in the southwest corner
located at (-71000 m, 58000 m). For the midwestern subdomain we use
the western-most 146 by 180 grid of cells when fitting
the statistical models. For the eastern subdomain we use the eastern-most
180 by 180 grid of cells and then omit 23 rows of cells in the north
and 17 rows of cells in the south, as these grid cells are outside
of the states containing data.

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in a grid cell or township.
\end{enumerate}
At a high level, the Bayesian statistical model estimates
composition across the domain, even in locations with sparse or no
data, by combining the raw composition data with the assumption that
composition varies in a smooth spatial fashion across the domain.
The information in the data is quantified by the data model, also
known as the likelihood. The assumption of smoothness is built into
the model by representing the true unknown spatially-varying composition
using a statistical spatial process representation that induces smoothing
of estimates across nearby locations. This spatial process representation
is a form of prior distribution and is a function of model parameters
called hyperparameters that determine the correlation structure of
the process and are also estimated based on the data.

The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
\emph{Data Product} section) and can be used in subsequent analyses.
The mean and standard deviation of the samples for each pair of cell
and taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

In the remainder of this section we provide the technical
specification of the model and of the computations involved in fitting
the model.

@@

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of \emph{P} Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the \emph{m} grid cells. In  \emph{Latent variable model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.
A multinomial probit model is similar to logistic
regression, used for modeling a binary outcome based on an underlying
probability the outcome will occur, but generalizes to modeling a
categorical variable based on probabilities for each category.

The critical component of the statistical model is the representation
of $\alpha_{p}(s)$ as a spatial process. A spatial
process is a statistical representation that models spatially-correlated
values. It provides a prior structure that serves to smooth across
noise in the observations and allows for prediction at locations based
on information from nearby locations, including interpolation to locations
with no data. Apart from the sum-to-one constraint, the taxa are considered
to be independent in the prior. We did not want to impose any structure
that ties the different taxa together, as any correlation will likely
vary across space.

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with \matern
covariance \cite{Lind:etal:2011}. These models are
specific statistical formulations of spatial processes that represent
spatial correlation by defining neighborhoods around each location
that are used to help inform predictions at the location.

@@

MRF models represent the neighborhood information
by working directly with the precision matrix (the
inverse of the covariance matrix) of the values of the spatial process,
so calculation of the prior density of $\alpha_{p}$ is computationally
simple \cite{Rue:Held:2005}. However, in situations
in which the likelihood is not normal, such
as our multinomial likelihood, it can be difficult to set up effective
MCMC algorithms that are able to move in the high-dimensional space
of $\alpha_{p}$. The latent variable representation
helps to alleviate this problem. Next we describe two alternative
spatial models that we considered; in \emph{Model comparison},
we evaluate the models on held-out data to choose
between the two.

@@

Our first model is a standard conditional autoregressive (CAR) model;
technical details can be found in \cite{Bane:etal:2004}. We use
a standard form of this model that treats the four
cardinal neighbors of each grid cell as the neighbors of the grid
cell. The corresponding precision matrix has diagonal elements, $Q_{ii}$,
equal to the number of neighbors for the \emph{i}th area (i.e., four except
for cells on the boundary of the domain), while $Q_{ik}=-1$ (the
negative of a weight of one) when areas \emph{i} and \emph{k} are neighbors
and $Q_{ik}=0$ when they are not. This gives the following model
for the values of $\alpha_{p}(s_{i})$ collected as a vector across
all of the grid cells, $i=1,\ldots,m$: 
\[
\alpha_{p}\sim\mbox{N}(0,\sigma_{p}^{2}Q^{-}).
\]
The use of the generalized inverse notation indicates that \emph{Q} is
not full-rank, but is of rank $m-1$; this gives an improper prior
on an implicit overall mean for the process values. Note
that we specify an explicit mean of zero because a non-zero mean would
not be identifiable in light of the implicit mean. This specification
is called an \textit{intrinsic conditional autoregression (ICAR)}
and we can write $Q=D-C$ where \emph{C} is the $m\times m$ adjacency
matrix defining the neighborhood relation of the locations; that is
$C{}_{ik}=1$ if locations \emph{i} and \emph{k} are neighbors and zero otherwise.
The matrix $D$ is an $m\times m$ diagonal matrix containing the
row sums of matrix $C$ as the diagonal entries, $D{}_{ii}={\displaystyle \sum_{k=1}^{m}C{}_{ik}}.$

@@

Gaussian processes (GP) are also standard models for spatial processes \cite{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of computational manipulations involving large covariance
matrices. Given this, \cite{Lind:etal:2011} proposed a new framework
for using Gaussian MRFs (GMRFs) as approximations to GPs, based on
the use of stochastic partial differential equations (SPDEs).

Gaussian processes are generally constructed using
one of a number of correlation functions that define how the strength
of correlation between the values of the process at two locations
decays as a function of the distance between the locations. We consider
Gaussian processes in the commonly-used \matern
class, using the following parameterization of the \matern correlation
function, 
\begin{equation}
R(d)=\frac{1}{\Gamma(\nu)2^{\nu-1}}\left(\frac{2\sqrt{\nu}d}{\rho}\right)^{\nu}\mathcal{K}_{\nu}\left(\frac{2\sqrt{\nu}d}{\rho}\right),\label{eq:Matern}
\end{equation}
where $d$ is Euclidean distance, $\rho$ is the spatial range parameter,
and $\mathcal{K}_{\nu}(\cdot)$ is the modified Bessel function of
the second kind, whose order is the smoothness (differentiability)
parameter, $\nu>0$. $\nu=0.5$ gives the exponential covariance.
For any pair of locations, $R(d)$ defines the correlation of the
process, (i.e., $\alpha_{p}(s)$ in our context), as a function of
the distance between the locations. Considering all pairs of locations,
this defines a correlation matrix for all locations of interest.

The approach of \cite{Lind:etal:2011} allows us to consider MRF
approximations to the \matern-based GP for $\nu=1$ and $\nu=2$.
Our second spatial model is this Lindgren approximation for \matern-based
GPs with $\nu=1$. To implement this Lindgren model, one modifies
the \emph{Q} matrix defined previously as follows based
on the technical specification of the precision matrix provided in
\cite{Lind:etal:2011}. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of \emph{Q} are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

The primary difference between the CAR and Lindgren models is that
the Lindgren model provides an additional degree of freedom by estimating
$\rho$. In particular $\rho$ allows us to estimate the locality
of the spatial smoothing. As $\rho$ decreases, the
model uses increasingly localized data to estimate the compositional
proportions at a given location, effectively averaging the empirical
proportions over smaller neighborhoods. In general, the \cite{Lind:etal:2011}
model will generally provide for a smoother estimate than the CAR
model \cite{Paci:2013}.

To ensure that the $\sigma^{2}$ parameter is mathematically equivalent
between the two models, we reparameterize, producing our second model:
\[
\alpha_{p}\sim\mbox{N}\left(\mu_{p},\sigma_{p}^{2}\cdot\frac{4\pi}{\rho_{p}^{2}}Q(\rho_{p})^{-1}\right)
\]

@@

Here we summarize the results of our analyses that inform the choice
between the CAR and SPDE models.
